Understanding the Landscape of Childlessness in the U.S.

### 2024-03-18 outputs/paper/paper.qmd
re-format paper

# Introduction {#sec-intro}
The American family landscape is undergoing a significant transformation, marked by a growing trend of childlessness. Traditionally viewed as a societal norm, having children now becomes a choice [@voluntaryChild]. We examine factors such as age, gender, education, socioeconomic class, health conditions and marital status, and how they have affected childbearing rates over the past 50 years.

These helps us understand not only the prevalence of childlessness within different segments of the population but also the potential drivers behind this trend. Existing research suggests a complex interplay of factors, including economic considerations, rising educational attainment, changing social norms related to career and family life, and individual preferences [@voluntaryChild].

We explore the impact of these factors on the rise of childlessness in America. By analyzing the data, we seek to answer crucial questions: What are some important factors contributing to this trend? How does childlessness manifest differently across demographic groups? Finally, what are the potential social, economic, and individual consequences of this new family structure in the United States? Understanding these dynamics is crucial in having an informed policy discussion and navigating the implications of childlessness in a rapidly changing world.

Our estimand is the relationship between age, gender, socioeconomic class, educational attainment, health status, marital status and the rate of childlessness in the United States from 1972 to 2022. Through our analysis, we found that the childless rate is consistently on the rise from 1972 to 2022, with a correlation coefficient of 0.73 between the childless percentage and time (or a 6.99% increase from 1972 to 2022). Factors that directly contribute to this rise are higher educational attainment, higher socioeconomic class, lower gender disparity, better health status and dissolution of marriages through divorces, separation and being widowed.

@@

## Exploring Gender Differences
It is crucial to acknowledge potential gender disparities in childbearing experiences. A separate analysis examining childlessness by gender (presented in @tbl-nochild-sex) aims to provide a comprehensive understanding by considering the unique experiences of men and women regarding childlessness. Data on gender is included from NORC [@gss] to allow for this investigation.

@@

## Exploring Socioeconomic Class and Educational Attainment
To gain a deeper understanding of the factors influencing these trends, we look into additional data sets from NORC [@gss] examining the relationship between childlessness and socioeconomic class (@tbl-nochild-class) and educational attainment (@tbl-nochild-degree).

@@

In addition, additional data sets from NORC [@gss] were also leveraged to understand the context of childlessness in America from 1972-2022. @tbl-nochild-health and @tbl-nochild-marital present an interesting connection between childlessness, health conditions and marital status.

@@

### Overview of Health Status's Impact on Childlessness

@@

```{r}
#| echo: false
#| include: true
#| warning: false
#| label: tbl-nochild-marital

@@

## Societal Implications of Rising Childlessness

@@

We are conducting a study to investigate the reasons behind childlessness in different groups of people. To do so, we will analyze various data sets, including information on age, gender, education, socioeconomic status, health conditions, and marital status. By examining these factors, we hope to gain insight into how they impact rates of childbearing.

>>>>

# Introduction {#sec-intro}

The American family landscape is undergoing a significant transformation, marked by a growing trend of childlessness. Traditionally viewed as a social norm, having children now becomes a choice [@voluntaryChild]. We examine factors such as age, gender, education, socioeconomic class, health conditions and marital status, and how they affect childbearing rates from 1972 to 2022. Existing research suggests a complex interplay of factors that contribute to childlessness, including economic considerations, rising educational attainment, shifting social norms related to career and family life, and individual preferences [@voluntaryChild].

By analyzing the data, we seek to answer crucial questions: What are some important factors contributing to this trend? How does childlessness manifest differently across demographic groups? Finally, what are the potential social, economic, and individual consequences of this new family structure in the United States? Understanding these dynamics is crucial in having an informed policy discussion and navigating the implications of childlessness in a rapidly changing world.

Thus, our estimand is the relationship between age, gender, socio-economic class, educational attainment, health status, marital status and the rate of childlessness in the United States from 1972 to 2022. Through our analysis, we found that childless rate is consistently on the rise from 1972 to 2022, with correlation coefficient of 0.73 between childless percentage and time (or a 6.99% increase over the last 50 years). Factors that directly contribute to this rise are higher educational attainment, higher socio-economic class, lower gender disparity, better health status and dissolution of marriages through divorces, separation and being widowed. In addition, global events like pandemics and recession also heighten the rate of childlessness from 1972-2022, including the AIDs pandemic, 9/11, 2008 Recession and the COVID-19 pandemic. During the 2008 Recession, from 2008 to 2014, the rate of childlessness increases the most, at 3%.

@@

## Exploring Gender Differences
A separate analysis examining childlessness by gender (presented in @tbl-nochild-sex) aims to provide a comprehensive understanding by considering the unique experiences of men and women regarding childlessness. Data on gender is included from NORC [@gss] to allow for this investigation.

@@

## Exploring Socioeconomic Class and Educational Attainment
We also look into additional data sets from NORC [@gss] examining the relationship between childlessness and socioeconomic class (@tbl-nochild-class) and educational attainment (@tbl-nochild-degree).

@@

We also use additional data sets from NORC [@gss] to understand the context of childlessness in America from 1972-2022. @tbl-nochild-health and @tbl-nochild-marital present an interesting connection between childlessness, health conditions and marital status.

@@

\newpage
### Overview of Health Status's Impact on Childlessness

@@

\newpage
```{r}
#| echo: false
#| include: true
#| warning: false
#| label: tbl-nochild-marital

@@

## Social and Ethical Implications of Rising Childlessness

@@

We are conducting a study to investigate the reasons behind childlessness in different groups of people. To do so, we will expand upon the GSS's existing questions and various data sets, including information on age, gender, education, socioeconomic status, health conditions, and marital status. By adding additional questions on these factors, we hope to gain more insight into how they impact childbearing rates.
